Trends from 30-Year Observations of Downward Solar Irradiance in Thessaloniki, Greece

### 2024-01-01 MDPI_submission.tex
Commit 2024-01-01 19:12

In order to unveil further the features of the variability of the three
datasets, Figure~\ref{fig:cusumnotrendmonthly} presents another set of
CUSUM plots using anomalies after the long-term linear trend is removed.
With this approach, periods when the CUSUMs diverge from zero can be
interpreted as a systematic variation of SDR from the climatological
mean. When the CUSUM is increasing, the anomalies values are above the
climatology and vice versa. Overall, for all- and cloudy-sky conditions
(Figures~\ref{fig:cusumnotrendmonthly-1}
and~\ref{fig:cusumnotrendmonthly-3}) we observe periods with anomalies
diverging from the climatological values, each lasting for several
years. These fluctuations are probably within the natural variability,
and no distinct changes are identified. The pattern in both datasets is
similar, suggesting prevalence in cloudy skies over Thessaloniki. For
clear skies (Figure~\ref{fig:cusumnotrendmonthly-2}) the distinct change
in 2004 is now clearer. The most likely reason for this change is the
monotonic reduction of aerosols in Thessaloniki. In that year, there was
a change in the rate of decrease in aerosol optical depth, as
illustrated in Figure 7 of \citet{Siomos2020}. This abrupt change in
CUSUMs lasted until about 2010, when the anomalies become again
variable.

\begin{figure}[h!]
    \begin{adjustwidth}{-\extralength}{0cm}
        {\centering 
            % \subfloat[All skies.\label{fig:cusumnotrendmonthly-1}]
                {\includegraphics[width=.32\linewidth]{./images/CumulativeMonthlyCuSumNOtrend-1} }\hfill
            % \subfloat[Clear skies.\label{fig:cusumnotrendmonthly-2}]
                {\includegraphics[width=.32\linewidth]{./images/CumulativeMonthlyCuSumNOtrend-5} }\hfill
            % \subfloat[Cloudy skies.\label{fig:cusumnotrendmonthly-3}]
                {\includegraphics[width=.32\linewidth]{./images/CumulativeMonthlyCuSumNOtrend-9} }
        }
        \caption{Cumulative sum plots of monthly SDR anomalies in (\%) for different sky conditions after removing the long-term linear trend: (a) all skies; (b) clear skies; (c) cloudy skies.}\label{fig:cusumnotrendmonthly}
\end{adjustwidth}
\end{figure}

@@

The solar zenith angle is a major factor affecting the SDR, since
increases in SZA leads to enhancement of the radiation path in the
atmosphere, especially in urban environments with human activities
emitting aerosols \citep{Wang2021}. In order to estimate the effect of
SZA on the SDR trends, we grouped the data in bins of \(1^\circ\) SZA,
and calculated the overall trend for each bin, separately for the daily
periods before noon and after noon (Figure~\ref{fig:szatrends}).
Although there are seasonal dependencies of the minimum SZA (see
Appendix, Figure~\ref{fig:SZAtrendSeason}), these dependencies are not
discussed further.

For all-sky conditions the brightening effect of SDR (positive trend)
increases with SZAs (Figures~\ref{fig:szatrends-1}) ranging from about
\(0.1\,\%/y\) to about \(0.7\,\%/y\) for the statistically significant
trends. The trends in the morning and afternoon hours are more or less
consistent with small differences at small SZAs which can be attributed
to effects on clear sky SDR from systematic diurnal patterns of aerosols
during the warm period of the year, consistently with the results
reported for China by \citet{Wang2021}. Note that SZAs less than
\(25^\circ\) can only occur during the warm period of the year around
noon when clear skies are more frequent. The increasing trend with SZA
is likely caused by the increased attenuation of SDR with SZA. The
effect is larger when aerosol and/or cloud layers are optically thicker,
therefore, decreases in aerosol and clouds through the study period will
result in larger positive trends of SDR at larger SZAs.

Under clear skies (Figure~\ref{fig:szatrends-2}), the trends are smaller
and less variable, ranging between \(0.1\) and \(0.15\,\%/y\) up to
\(77^\circ\) SZA. At higher SZAs and in the afternoon hours, there is a
sharp increase in the trend up to \(0.3\,\%/y\), which may have been
caused by the long path length of radiation through the atmosphere as
discussed above for the all-sky conditions. The small differences in the
trend between morning and afternoon between \(35^\circ\) and
\(60^\circ\) SZA is likely a result of less attenuation of SDR in the
morning hours due to lesser amounts of aerosols and a shallower boundary
layer.

For cloudy-sky conditions (Figure~\ref{fig:szatrends-3}), we cannot
discern any significant dependence of the SDR trend with SZA as the
variability of irradiance is dominated by the cloud effects leading to
insignificant trends. Statistically significant trends appear only in
the afternoon and for SZAs larger than \(60^\circ\). The sharp increase
of the trend at SZAs larger than \(\sim{75}^{\circ}\), observed also for
clear skies, is probably associated with stronger attenuation by clouds
under oblique incidence angles, which result also in smaller
variability.

\begin{figure}[h!]
    \begin{adjustwidth}{-\extralength}{0cm}
        {\centering 
            \subfloat[All skies.\label{fig:szatrends-1}]
                {\includegraphics[width=.32\linewidth]{./images/SzaTrends-1}}\hfill
            \subfloat[Clear skies.\label{fig:szatrends-2}]
                {\includegraphics[width=.32\linewidth]{./images/SzaTrends-4}}\hfill
            \subfloat[Cloudy skies.\label{fig:szatrends-3}]
                {\includegraphics[width=.32\linewidth]{./images/SzaTrends-7}}
        }
        \caption{Long term trends of daily SDR as a function of SZA for (a) all-sky, (b) clear-sky and (c) cloudy-sky conditions, separately for morning and afternoon periods. Solid shapes represent statistically significant trends ($p < 0.005$).}\label{fig:szatrends}
    \end{adjustwidth}
\end{figure}

\hypertarget{long-term-sdr-trends-by-season}{%
\subsection{Long term SDR trends by
season}\label{long-term-sdr-trends-by-season}}

Similarly to the long term trends from daily means of SDR discussed
above, we have calculated the trend for the three sky conditions and for
each season of the year, using the corresponding mean monthly anomalies
(Figure~\ref{fig:seasonalALL} and Table~\ref{tab:trendseasontable}).
Table~\ref{tab:trendseasontable} contains also the \(2\sigma\) standard
error, the Pearson's correlation coefficient R and the corresponding
p-value. The winter linear trends exhibit generally the largest R values
ranging between \(0.54\) and \(0.60\,\%/y\). For all-sky conditions the
trend in SDR in winter is the largest (\(0.7\,\%/y\)), followed by the
trend in autumn (\(0.42\,\%/y\), a value close to the long term trend)
both statistically significant at the \(95\,\%\) confidence level. In
spring and summer, the trends are much smaller and of lesser statistical
significance. These seasonal differences indicate a possible relation of
the trends in SDR to trends in clouds during winter and autumn. For
clear-skies, the trend in winter is \(0.4\,\%/y\) and is associated with
the decreasing trend in aerosol optical depth \citep{Siomos2020}.
Moreover, it is almost half of that for all-skies, which is another
indication of a decreasing trend in cloud optical thickness. In other
seasons the clear-sky trend is very small (below \(0.1\,\%/y\)).
Finally, for cloudy-skies the winter trend is the largest
(\(0.76\,\%/y\)) and greater than for all-skies, followed by a much
smaller trend in autumn (\(0.19\,\%/y\)).

@@

@@

>>>>

In order to unveil further the features of the variability of the three
datasets, Figure~\ref{fig:cusumnotrendmonthly} presents another set of
CUSUM plots using anomalies after the long-term linear trend is removed.
With this approach, periods when the CUSUMs diverge from zero can be
interpreted as a systematic variation of SDR from the climatological
mean. When the CUSUM is increasing, the anomalies values are above the
climatology, and vice versa. Overall, for all- and cloudy-sky conditions
(Figure~\ref{fig:cusumnotrendmonthly}a,c), we observe periods with
anomalies diverging from the climatological values, each lasting for
several years. These fluctuations are probably within the natural
variability, and no distinct changes are identified. The pattern in both
datasets is similar, suggesting prevalence in cloudy skies over
Thessaloniki. For clear skies (Figure~\ref{fig:cusumnotrendmonthly}b),
the distinct change in 2004 is now clearer. The most likely reason for
this change is the monotonic reduction of aerosols in Thessaloniki. In
that year, there was a change in the rate of decrease in aerosol optical
depth, as illustrated in Figure 7 of \citet{Siomos2020}. This abrupt
change in CUSUMs lasted until about 2010, when the anomalies become
again variable.

\begin{figure}[h!]
    \begin{adjustwidth}{-\extralength}{0cm}
        {\centering 
            % \subfloat[All skies.\label{fig:cusumnotrendmonthly-1}]
                {\includegraphics[width=.32\linewidth]{./images/CumulativeMonthlyCuSumNOtrend-1} }\hfill
            % \subfloat[Clear skies.\label{fig:cusumnotrendmonthly-2}]
                {\includegraphics[width=.32\linewidth]{./images/CumulativeMonthlyCuSumNOtrend-5} }\hfill
            % \subfloat[Cloudy skies.\label{fig:cusumnotrendmonthly-3}]
                {\includegraphics[width=.32\linewidth]{./images/CumulativeMonthlyCuSumNOtrend-9} }
        }
    \end{adjustwidth}
    \caption{Cumulative sum plots of monthly SDR anomalies in (\%) for different sky conditions after removing the long-term linear trend: (a) all skies; (b) clear skies; (c) cloudy skies.}\label{fig:cusumnotrendmonthly}
\end{figure}

@@

The solar zenith angle is a major factor affecting the SDR, as increases
in SZA leads to enhancement of the radiation path in the atmosphere,
especially in urban environments with human activities emitting aerosols
\citep{Wang2021}. In order to estimate the effect of SZA on the SDR
trends, we grouped the data in bins of \(1^\circ\) SZA and calculated
the overall trend for each bin separately for the daily periods before
noon and after noon (Figure~\ref{fig:szatrends}). Although there are
seasonal dependencies of the minimum SZA (see Appendix \ref{appendix},
Figure~\ref{fig:SZAtrendSeason}), these dependencies are not discussed
further.

For all-sky conditions, the brightening effect of SDR (positive trend)
increases with SZAs (Figure~\ref{fig:szatrends}a), ranging from about
\(0.1\,\%/y\) to about \(0.7\,\%/y\) for the statistically significant
trends. The trends in the morning and afternoon hours are more or less
consistent with small differences at small SZAs, which can be attributed
to effects on clear sky SDR from systematic diurnal patterns of aerosols
during the warm period of the year, consistent with the results reported
for China by \citet{Wang2021}. Note that SZAs less than \(25^\circ\) can
only occur during the warm period of the year around noon when clear
skies are more frequent. The increasing trend with SZA is likely caused
by the increased attenuation of SDR with SZA. The effect is larger when
aerosol and/or cloud layers are optically thicker; therefore, decreases
in aerosol and clouds through the study period will result in larger
positive trends of SDR at larger SZAs.

Under clear skies (Figure~\ref{fig:szatrends}b), the trends are smaller
and less variable, ranging between \(0.1\) and \(0.15\,\%/y\) up to
\(77^\circ\) SZA. At higher SZAs and in the afternoon hours, there is a
sharp increase in the trend up to \(0.3\,\%/y\), which may have been
caused by the long path length of radiation through the atmosphere as
discussed above for the all-sky conditions. The small differences in the
trend between morning and afternoon between \(35^\circ\) and
\(60^\circ\) SZA is likely a result of less attenuation of SDR in the
morning hours due to lesser amounts of aerosols and a shallower boundary
layer.

For cloudy-sky conditions (Figure~\ref{fig:szatrends}c), we cannot
discern any significant dependence of the SDR trend with SZA, as the
variability of irradiance is dominated by the cloud effects leading to
insignificant trends. Statistically significant trends appear only in
the afternoon and for SZAs larger than \(60^\circ\). The sharp increase
of the trend at SZAs larger than \(\sim{75}^{\circ}\), observed also for
clear skies, is probably associated with stronger attenuation by clouds
under oblique incidence angles, which also result in smaller
variability.

\begin{figure}[h!]
    \begin{adjustwidth}{-\extralength}{0cm}
        {\centering 
            % \subfloat[All skies.\label{fig:szatrends-1}]
                {\includegraphics[width=.32\linewidth]{./images/SzaTrends-1}}\hfill
            % \subfloat[Clear skies.\label{fig:szatrends-2}]
                {\includegraphics[width=.32\linewidth]{./images/SzaTrends-4}}\hfill
            % \subfloat[Cloudy skies.\label{fig:szatrends-3}]
                {\includegraphics[width=.32\linewidth]{./images/SzaTrends-7}}
        }
    \end{adjustwidth}
    \caption{Long-term trends of daily SDR as a function of SZA for (a) all-sky, (b) clear-sky and (c) cloudy-sky conditions, separately for morning and afternoon periods. Solid shapes represent statistically significant trends ($p < 0.005$).}\label{fig:szatrends}
\end{figure}

\hypertarget{long-term-sdr-trends-by-season}{%
\subsection{Long-Term SDR Trends by
Season}\label{long-term-sdr-trends-by-season}}

Similarly to the long term trends from daily means of SDR discussed
above, we have calculated the trend for the three sky conditions and for
each season of the year, using the corresponding mean monthly anomalies
(Figure~\ref{fig:seasonalALL} and Table~\ref{tab:trendseasontable}).
Table~\ref{tab:trendseasontable} also contains the \(2\sigma\) standard
error, the Pearson's correlation coefficient R and the corresponding
p-value. The winter linear trends exhibit generally the largest R values
ranging between \(0.54\) and \(0.60\,\%/y\). For all-sky conditions, the
trend in SDR in winter is the largest (\(0.7\,\%/y\)), followed by the
trend in autumn (\(0.42\,\%/y\), a value close to the long term trend)
both statistically significant at the \(95\,\%\) confidence level. In
spring and summer, the trends are much smaller and of lesser statistical
significance. These seasonal differences indicate a possible relation of
the trends in SDR to trends in clouds during winter and autumn. For
clear-skies, the trend in winter is \(0.4\,\%/y\) and is associated with
the decreasing trend in aerosol optical depth \citep{Siomos2020}.
Moreover, it is almost half of that for all-skies, which is another
indication of a decreasing trend in cloud optical thickness. In other
seasons the clear-sky trend is very small (below \(0.1\,\%/y\)).
Finally, for cloudy-skies the winter trend is the largest
(\(0.76\,\%/y\)) and greater than for all-skies, followed by a much
smaller trend in autumn (\(0.19\,\%/y\)).

@@

\institutionalreview{Not applicable.}

\informedconsent{Not applicable.}

@@

\conflictsofinterest{The authors declare no conflict of interest.}
